---
title: "Descriptive Statistics: PTLD Clinical Phenotypes"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
author: "David Arthur"
output:
  bookdown::word_document2:
    number_sections: false
  word_document: default
  html_document:
    df_print: paged
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../Results",  
    )
  })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      ft.align = "left", ft.keepnext = TRUE, cache = FALSE)
```

```{r libraries, include = FALSE}

# Install required packages if not already installed
# install.packages(c("haven", "dplyr", "tidyr", "gtsummary", "naniar", "ggplot2"))

library(haven)
library(dplyr)
library(tidyr)
library(gtsummary)
library(naniar)
library(ggplot2)
library(labelled)
library(flextable)
library(officer)
library(tidyverse)

```


```{r}

theme_gtsummary_compact()

```


# Introduction

This report provides initial descriptive statistics and missing data analysis for the IeDEA TB-SRN project: **Clinical Phenotypes of Post-Tuberculosis Lung Disease**. 

The analysis is based on two SAS datasets: `MR264DEMOG.SAS7BDAT` (baseline demographics and TB treatment outcomes) and `MR264VISITS.SAS7BDAT` (longitudinal clinical and psychosocial data).

# Data Import and Preparation

```{r set-paths, include = FALSE}

current.user <- Sys.getenv("USERNAME")
project.path <- paste0('C:/Users/', current.user, '/Box/PROJECT 00113062-2□ K23 Cluster Analysis')
data_path <- paste0(project.path)

```

```{r, eval = FALSE}

# Look at the raw integer levels for the categorical variables
vars_to_check <- c(
  "previous_tb_treated", "obstructive_pattern", "restrictive_pattern", 
  "cavitation", "hivstatus", "sex_at_birth", "tb_category", 
  "tb_tx_outcomes", "Smoker", "death"
)

# This will print the unique values found in the raw demographics dataset
lapply(df_demog[names(df_demog) %in% vars_to_check], function(x) sort(unique(na.omit(x))))

# And for the visits dataset
lapply(df_visits[names(df_visits) %in% vars_to_check], function(x) sort(unique(na.omit(x))))

```


```{r load_data}
# Define paths to your SAS datasets
path_demog <- paste0(data_path, "/MR264DEMOG.SAS7BDAT")
path_visits <- paste0(data_path, "/MR264VISITS.SAS7BDAT")

# Import datasets
# Note: haven::read_sas brings in SAS labels and formats. 
df_demog <- read_sas(path_demog)
df_visits <- read_sas(path_visits)
# 1. Clean Demographics Dataset
df_demog_clean <- df_demog %>%
  mutate(
    sex_at_birth = factor(sex_at_birth, levels = c(1, 2, 88), labels = c("Male", "Female", "Other")),
    hivstatus = factor(hivstatus, levels = c(0, 1), labels = c("Negative", "Positive")),
    tb_category = factor(tb_category, levels = c(1, 2, 3, 4, 5, 88), labels = c("New case", "Relapse", "Treatment after failure", "Treatment after loss to follow-up", "Transfer in", "Other")),
    resistance_pattern = factor(resistance_pattern, levels = c(1, 2), labels = c("Drug-susceptible (DS-TB)", "Drug resistant (DR-TB)")),
    tb_tx_outcomes = factor(tb_tx_outcomes, levels = 1:6, labels = c("Cure", "Treatment Completed", "Treatment failed", "Died (any cause)", "Lost to follow-up", "Transferred out from study site")),
    dwelling_location = factor(dwelling_location, levels = c(1, 2, 3), labels = c("City/urban area", "Peri-urban area", "Rural area")),
    previous_tb_treated = factor(previous_tb_treated, levels = c(0, 1), labels = c("No", "Yes")),
    across(c(asthma_hx, copd_hx, diabetes_hx, death), ~case_when(
      . == 1 ~ "Yes",
      . %in% c(0, 2) ~ "No",
      TRUE ~ NA_character_
    )),
    region = factor(
      region, 
      levels = c("CA", "CN", "EA", "WA"), 
      labels = c("Central Africa", "Central/South America", "East Africa", "West Africa")
    )
  )

# 2. Clean Visits Dataset
df_visits_clean <- df_visits %>%
  mutate(
    redcap_event_name = factor(
      redcap_event_name,
      levels = c(
        "multivisit_forms_arm_1",
        "baseline_arm_1",
        "month_1_arm_1",
        "month_2_arm_1",
        "end_of_tx_arm_1",
        "6m_posttx_arm_1",
        "12m_posttx_arm_1",
        "18m_posttx_arm_1",
        "24m_posttx_arm_1",
        "tx_frw_arm_1"
      ),
      labels = c(
        "Multi-Visit-Forms",
        "Baseline",
        "Month 1",
        "Month 2",
        "End of TX",
        "6-M Post-TX",
        "12-M Post-TX",
        "18-M Post-TX",
        "24-M Post-TX",
        "Tx F/R/W"
      )
    ),
    region = factor(
      region, 
      levels = c("CA", "CN", "EA", "WA"), 
      labels = c("Central Africa", "Central/South America", "East Africa", "West Africa")
    )
  ) %>%
  mutate(
    AnyRespSymptoms = factor(AnyRespSymptoms, levels = c(0, 1), labels = c("No", "Yes")),
    PersistentRespSymptoms = factor(PersistentRespSymptoms, levels = c(0, 1), labels = c("No", "Yes")),
    hypoxemia = factor(hypoxemia, levels = c(0, 1), labels = c("No", "Yes")),
    obstructive_pattern = factor(obstructive_pattern, levels = c(1, 2), labels = c("Yes", "No")),
    restrictive_pattern = factor(restrictive_pattern, levels = c(1, 2), labels = c("Yes", "No")),
    complete_spirometry = factor(complete_spirometry, levels = c(1, 2, 3), labels = c("Yes", "No", "Not applicable")),
    Smoker = factor(Smoker, levels = c(0, 1, 2), labels = c("Never smoker", "Current smoker", "Past smoker")),
    tobacco_hx = factor(tobacco_hx, levels = c(0, 1, 2), labels = c("Never Used Tobacco", "Current Tobacco User", "Past Tobacco User")),
    missed_dose_last_thirty = factor(missed_dose_last_thirty, levels = c(0, 1), labels = c("No", "Yes")),
    packsperday = factor(packsperday, levels = 1:5, labels = c("<=1/2", ">1/2 - 1", "1 - 1.5", ">1.5 - 2", ">2")),
    S2SExertHypoxemia = factor(S2SExertHypoxemia, levels = c(0, 1), labels = c("No", "Yes"))
  ) %>%
  mutate(across(c(cavitation, nodules_masses, alveolar_opacity, pleural_effusion), 
                ~factor(., levels = c(1, 2), labels = c("Yes", "No"))))

# 2b. Clean 6-month Visits Dataset
df_6m_visit <- df_visits_clean %>%
  filter(redcap_event_name == "6-M Post-TX")
  

# 3. Pull End-of-Treatment (EOT) Imaging variables for clustering
# Prefix them with 'eot_' to prevent join collisions and clarify the timepoint
df_eot_imaging <- df_visits_clean %>%
  filter(redcap_event_name == "End of TX") %>%
  select(
    patient, 
    eot_cavitation = cavitation, 
    eot_nodules_masses = nodules_masses, 
    eot_alveolar_opacity = alveolar_opacity, 
    eot_pleural_effusion = pleural_effusion
  )

# 4. Master Join and PTLD Classification (Updated for Strict Missingness)
df_master <- df_demog_clean %>%
  inner_join(df_6m_visit, by = c("patient", "region")) %>%
  left_join(df_eot_imaging, by = "patient") %>%
  mutate(
    # PTLD Antecedent Requirements
    successful_tx = if_else(tb_tx_outcomes %in% c("Cure", "Treatment Completed"), "Yes", "No"),
    no_active_tb = if_else(is.na(RecurrenceAfterEOT) | RecurrenceAfterEOT != 1, "Yes", "No"),
    no_tx_fail = if_else(is.na(TxFailAfterEOT) | TxFailAfterEOT != 1, "Yes", "No"),
    antecedent_met = if_else(successful_tx == "Yes" & no_active_tb == "Yes" & no_tx_fail == "Yes", "Yes", "No"),

    # PTLD Domains - Explicitly handling NAs
    domain_symptoms = case_when(
      PersistentRespSymptoms == "Yes" ~ 1,
      PersistentRespSymptoms == "No" ~ 0,
      TRUE ~ NA_real_
    ),
    
    domain_lung_function = case_when(
      obstructive_pattern == "Yes" | restrictive_pattern == "Yes" ~ 1,
      obstructive_pattern == "No" & restrictive_pattern == "No" ~ 0,
      TRUE ~ NA_real_
    ),
    
    domain_imaging = 0, # Not collected at 6M, so cohort baseline for this domain is 0
    
    # PTLD Classification (Strict Missingness Handling)
    ptld_classification = case_when(
      antecedent_met == "No" ~ "Excluded: Antecedent criteria not met",
      
      # Patient has valid data for BOTH domains and qualifies for Disease (2 domains)
      domain_symptoms + domain_lung_function >= 2 ~ "Post-tuberculosis lung disease",
      
      # Patient has valid data for BOTH domains and qualifies for Abnormality (1 domain)
      domain_symptoms + domain_lung_function == 1 ~ "Post-tuberculosis lung abnormality",
      
      # Patient has valid data for BOTH domains and is completely normal (0 domains)
      domain_symptoms == 0 & domain_lung_function == 0 ~ "No PTLD or abnormality",
      
      # If any of the above conditions fail to evaluate to TRUE (i.e., because of an NA), 
      # they are routed here to prevent misclassification.
      TRUE ~ NA_character_
    ),
    
    # Subcategories
    ptld_subcategory = case_when(
      ptld_classification == "Post-tuberculosis lung disease" ~ "Lung function and symptoms without imaging",
      ptld_classification == "Post-tuberculosis lung abnormality" & domain_lung_function == 1 ~ "Only lung function",
      ptld_classification == "Post-tuberculosis lung abnormality" & domain_symptoms == 1 ~ "Only symptoms",
      TRUE ~ ptld_classification # Passes through "Unknown/Missing Data" or "Excluded"
    )
  )

# 5. Apply Variable Labels (Labelled Package)
df_master <- df_master %>%
  set_variable_labels(
    # Demographics & Clinical Base
    region = "Study Region",
    current_age = "Age (years)",
    sex_at_birth = "Sex at Birth",
    hivstatus = "HIV Status",
    bmi = "Body Mass Index",
    bsa = "Body Surface Area",
    tb_category = "Previous TB Classification",
    resistance_pattern = "Drug Susceptibility",
    asthma_hx = "History of Asthma",
    copd_hx = "History of COPD",
    diabetes_hx = "History of Diabetes",
    tb_tx_outcomes = "TB Treatment Outcomes",
    dwelling_location = "Dwelling Location",
    previous_tb_treated = "Previous TB Treated",
    
    # 6-Month Outcomes (Impact)
    SGRQ_TOTAL_SCORE = "SGRQ Total Score",
    phq9_total = "PHQ-9 Total Score (Depression)",
    min_completion_rate = "Sit-to-Stand: Repetitions (1 min)",
    S2SExertHypoxemia = "Sit-to-Stand: Exertional Hypoxemia",
    
    # PTLD Definition
    complete_spirometry = "Lung Function Tests Completed",
    obstructive_pattern = "Obstructive Spirometry Pattern",
    restrictive_pattern = "Restrictive Spirometry Pattern",
    hypoxemia = "Resting Hypoxemia (SpO2 < 90%)",
    AnyRespSymptoms = "Any Respiratory Symptoms",
    PersistentRespSymptoms = "Chronic/Persistent Respiratory Symptoms",
    ptld_classification = "PTLD Consensus Classification",
    ptld_subcategory = "PTLD Sub-category (Domains)",
    
    # Clustering Features
    Smoker = "Current Smoking Status",
    tobacco_hx = "Tobacco Use History",
    packsperday = "Packs per Day",
    a1c_percent = "Hemoglobin A1c (%)",
    missed_dose_last_thirty = "Missed TB Dose (Last 30 Days)",
    fev1_liters = "FEV1 (Liters)",
    fvc_liters = "FVC (Liters)",
    eot_cavitation = "CXR (EOT): Cavitation",
    eot_nodules_masses = "CXR (EOT): Nodules or Masses",
    eot_alveolar_opacity = "CXR (EOT): Alveolar Opacity",
    eot_pleural_effusion = "CXR (EOT): Pleural Effusion"
  )
```


```{r, eval = FALSE}

df_visits_clean %>%
  tbl_summary(
    include = c(
      contains("smok"),
      missed_dose_last_thirty
    ),
    by = redcap_event_name
  )

df_master %>%
  tbl_cross(PersistentRespSymptoms, AnyRespSymptoms)

```

# Study Follow-up and Retention

```{r table_visit_retention_fixed}
# 1. Create a wide dataset: One row per patient in the baseline cohort
df_retention_wide <- df_demog_clean %>%
  select(patient) %>% # Establishes the fixed denominator (Total N)
  left_join(
    df_visits_clean %>%
      # Keep only patient and visit name, removing any duplicate rows per visit
      distinct(patient, redcap_event_name) %>%
      # Create a flag indicating the patient had a record for this visit
      mutate(attended = 1),
    by = "patient"
  ) %>%
  # Pivot so each visit becomes its own column
  pivot_wider(
    names_from = redcap_event_name,
    values_from = attended,
    values_fill = 0, # If they didn't have a record, fill with 0
    names_sort = TRUE # Forces columns to sort by factor levels 
  ) %>%
  # Convert 1/0 flags to Yes/No factors for clean table output
  mutate(across(-patient, ~factor(., levels = c(1, 0), labels = c("Yes", "No"))))

# 2. Generate the summary table
df_retention_wide %>%
  select(-patient) %>% # Remove the ID column so it isn't summarized
  tbl_summary(
    # Tell gtsummary to ONLY show the count/percentage of "Yes" for each column
    value = list(everything() ~ "Yes"), 
    missing = "no" # We explicitly handled missingness with the 0 fill above
  ) %>%
  modify_header(
    label = "**Study Visit Timepoint**",
    stat_0 = "**Patients Retained (N = {N})**"
  ) %>%
  modify_caption("**Patient Retention by Visit (Denominator = Full Baseline Cohort)**") %>%
  bold_labels()  %>%
  as_flex_table() %>% 
  flextable::set_caption('Follow-up retention by timepoint',
                         autonum = officer::run_autonum(seq_id = "tab",
                                                        bkm = "RetentionTimepoint")) %>%
  bold(bold = TRUE, part = "header") %>%
  autofit() %>%
  set_table_properties(layout = "autofit") %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 11, part = "all")

```


# Demographic and Clinical Characteristics

```{r table_demographics}
df_master %>%
  select(
    region, current_age, sex_at_birth, hivstatus, bmi, bsa,
    tb_category, resistance_pattern, asthma_hx, copd_hx, diabetes_hx,
    tb_tx_outcomes, previous_tb_treated, dwelling_location
  ) %>%
  tbl_summary(
    by = region,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany",
    missing_text = "Missing/Not Assessed"
  ) %>%
  add_overall() %>%
  modify_header(label = "**Demographic & Clinical Variable**") %>%
  bold_labels()  %>%
  as_flex_table() %>% 
  flextable::set_caption('Variables from demographic dataset by region',
                         autonum = officer::run_autonum(seq_id = "tab",
                                                        bkm = "VarsDemo")) %>%
  bold(bold = TRUE, part = "header") %>%
  autofit() %>%
  set_table_properties(layout = "autofit") %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 11, part = "all")
```



# PTLD Definition

## Meeting antecedent requirements

```{r}

# Create a summary table of the PTLD antecedent requirements
df_master %>%
  # Select the raw variables and the derived antecedent flags
  select(
    tb_tx_outcomes,
    successful_tx,
    RecurrenceAfterEOT,
    TxFailAfterEOT,
    no_active_tb,
    no_tx_fail,
    antecedent_met
  ) %>%
  # Temporarily format the raw EOT numeric flags as Yes/No for cleaner table output
  mutate(
    RecurrenceAfterEOT = factor(RecurrenceAfterEOT, levels = c(0, 1), labels = c("No", "Yes")),
    TxFailAfterEOT = factor(TxFailAfterEOT, levels = c(0, 1), labels = c("No", "Yes"))
  ) %>%
  # Apply clear labels for the summary table
  set_variable_labels(
    tb_tx_outcomes = "Raw TB Treatment Outcome",
    successful_tx = "Requirement 1: Successfully Treated (Cured/Completed)",
    RecurrenceAfterEOT = "Raw Flag: Recurrence after End of Treatment",
    TxFailAfterEOT = "Raw Flag: Treatment Failure after End of Treatment",
    no_active_tb = "Requirement 2a: No Active TB (No Recurrence)",
    no_tx_fail = "Requirement 2b: No Active TB (No Post-EOT Failure)",
    antecedent_met = "ALL Antecedent Requirements Met (Eligible for PTLD Evaluation)"
  ) %>%
  tbl_summary(
    statistic = list(all_categorical() ~ "{n} ({p}%)"),
    missing = "ifany",
    missing_text = "Missing/Unknown"
  ) %>%
  modify_header(label = "**Antecedent Requirement**") %>%
  bold_labels()  %>%
  as_flex_table() %>% 
  flextable::set_caption('PTLD antecedent requirements',
                         autonum = officer::run_autonum(seq_id = "tab",
                                                        bkm = "PTLDAnteced")) %>%
  bold(bold = TRUE, part = "header") %>%
  autofit() %>%
  set_table_properties(layout = "autofit") %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 11, part = "all")

```
\newpage

## PTLD Classification

By the definition currently applied in the table below, patients who meet one of the criteria but don't have data available for the other (e.g. they have respiratory symptoms but their spirometry measurement at 6 months is missing) are counted as "Missing", rather than "Post-tuberculosis lung abnormality". 

```{r table_ptld}
# Filter only to patients who meet antecedent requirements for accurate reporting
df_master %>%
  filter(antecedent_met == "Yes") %>%
  select(
    complete_spirometry, obstructive_pattern, restrictive_pattern,
    hypoxemia, AnyRespSymptoms, PersistentRespSymptoms,
    ptld_classification, ptld_subcategory
  ) %>%
  tbl_summary(
    statistic = list(all_categorical() ~ "{n} ({p}%)"),
    missing = "ifany",
    missing_text = "Missing/Not Assessed"
  ) %>%
  modify_header(label = "**Reporting Framework Variable**") %>%
  bold_labels() %>%
  as_flex_table() %>% 
  flextable::set_caption('PTLD Variables and Classification',
                         autonum = officer::run_autonum(seq_id = "tab",
                                                        bkm = "PTLDVarsClass")) %>%
  bold(bold = TRUE, part = "header") %>%
  autofit() %>%
  set_table_properties(layout = "autofit") %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 11, part = "all")
```

## Crosstabulation of PTLD domains

```{r, tab.cap = 'PTLD Domains'}

# df_master %>%
#   filter(antecedent_met == "Yes") %>%
#   tbl_cross(obstructive_pattern, restrictive_pattern)

df_master %>%
  filter(antecedent_met == "Yes") %>%
  tbl_cross(domain_lung_function, domain_symptoms) %>%
  bold_labels() %>%
  as_flex_table() %>% 
  flextable::set_caption('PTLD Domains',
                         autonum = officer::run_autonum(seq_id = "tab",
                                                        bkm = "PTLDDomains")) %>%
  bold(bold = TRUE, part = "header") %>%
  autofit() %>%
  set_table_properties(layout = "autofit") %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 11, part = "all")


# df_master %>%
#   filter(antecedent_met == "Yes") %>%
#   tbl_summary(
#     include = c(obstructive_pattern, restrictive_pattern),
#     by = region
#   )


```


# Other Features for Cluster Analysis

These variables capture physiological state, comorbidities, and imaging phenotypes that will serve as the input matrix for the unsupervised machine learning models.

```{r table_clustering}
df_master %>%
  select(
    Smoker, tobacco_hx, packsperday, a1c_percent, missed_dose_last_thirty,
    fev1_liters, fvc_liters, eot_cavitation, eot_nodules_masses, 
    eot_alveolar_opacity, eot_pleural_effusion
  ) %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany",
    missing_text = "Missing/Not Assessed"
  ) %>%
  modify_header(label = "**Clustering Feature**") %>%
  bold_labels()  %>%
  as_flex_table() %>% 
  flextable::set_caption('Potential variables for cluster analysis',
                         autonum = officer::run_autonum(seq_id = "tab",
                                                        bkm = "VarsCluster")) %>%
  bold(bold = TRUE, part = "header") %>%
  autofit() %>%
  set_table_properties(layout = "autofit") %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 11, part = "all")
```

\newpage

# Post-Treatment Outcomes

These variables measure the downstream functional and quality-of-life impact of the lung disease and will be held out of the cluster analysis to serve as evaluation endpoints.

```{r table_outcomes_site}
df_master %>%
  select(
    region, SGRQ_TOTAL_SCORE, phq9_total, min_completion_rate, S2SExertHypoxemia
  ) %>%
  tbl_summary(
    by = region,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany",
    missing_text = "Missing/Not Assessed"
  ) %>%
  modify_header(label = "**6-Month Outcome Variable**") %>%
  add_overall() %>%
  bold_labels() %>%
  as_flex_table() %>% 
  flextable::set_caption('6-month post-tx outcome variables by site',
                         autonum = officer::run_autonum(seq_id = "tab",
                                                        bkm = "OutcomesSite")) %>%
  bold(bold = TRUE, part = "header") %>%
  autofit() %>%
  set_table_properties(layout = "autofit") %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 11, part = "all")

```

&nbsp;<br>
&nbsp;<br>

```{r table_outcomes_timepoint}

df_visits_clean %>%
  mutate(redcap_event_name = fct_drop(redcap_event_name)) %>%
  select(
    redcap_event_name, SGRQ_TOTAL_SCORE, phq9_total, min_completion_rate, S2SExertHypoxemia
  ) %>%
  tbl_summary(
    by = redcap_event_name,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany",
    missing_text = "Missing/Not Assessed"
  ) %>%
  modify_header(label = "**Outcome Variable**") %>%
  add_overall() %>%
  bold_labels() %>%
  as_flex_table() %>% 
  flextable::set_caption('Outcome variables by timepoint',
                         autonum = officer::run_autonum(seq_id = "tab",
                                                        bkm = "OutcomesTimepoint")) %>%
  bold(bold = TRUE, part = "header") %>%
  autofit() %>%
  set_table_properties(layout = "autofit") %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 11, part = "all")

```

<!-- # Missing Data Analysis -->

<!-- ## Missingness in Core PTLD Variables -->

<!-- ```{r missing_ptld_vars} -->
<!-- df_master %>% -->
<!--   # Filter to those meeting antecedent requirements to see missingness in the eligible cohort -->
<!--   filter(antecedent_met == "Yes") %>% -->
<!--   select( -->
<!--     PersistentRespSymptoms, -->
<!--     complete_spirometry, -->
<!--     obstructive_pattern, -->
<!--     restrictive_pattern, -->
<!--     hypoxemia, -->
<!--     ptld_classification -->
<!--   ) %>% -->
<!--   gg_miss_var() +  -->
<!--   labs(title = "Missing Values in Core PTLD Definition Variables (Eligible Cohort)", -->
<!--        y = "Number of missing cases") + -->
<!--   theme_minimal() -->

<!-- ``` -->

<!-- ## Missingness Patterns -->

<!-- ```{r} -->

<!-- df_master %>% -->
<!--   select( -->
<!--     Smoker, tobacco_hx, packsperday, a1c_percent, missed_dose_last_thirty, -->
<!--     fev1_liters, fvc_liters, eot_cavitation, eot_nodules_masses,  -->
<!--     eot_alveolar_opacity, eot_pleural_effusion -->
<!--   ) %>% -->
<!--   vis_miss(warn_large_data = FALSE) + -->
<!--   theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9))  -->

<!-- ``` -->


<!-- ## Patterns of Co-occurring Missing Data at 6-Month Follow-up -->

<!-- ```{r, fig.cap = "Patterns of Co-occurring Missing Data at 6-Month Follow-up. Horizontal bars indicate total missing cases; connected nodes represent the frequency of missing data combinations within individual patients."} -->

<!-- df_master %>% -->
<!--   select( -->
<!--     complete_spirometry, -->
<!--     SGRQ_TOTAL_SCORE, -->
<!--     phq9_total, -->
<!--     min_completion_rate, -->
<!--     PersistentRespSymptoms -->
<!--   ) %>% -->
<!--   gg_miss_upset( -->
<!--     nsets = 5,           # Number of variables to include in the plot -->
<!--     nintersects = 10,    # Top 10 most common combinations of missingness -->
<!--     text.scale = 1.2     # Adjust text scale for readability -->
<!--   ) -->

<!-- ``` -->



<!-- <!-- # Match patients in the datasets --> -->

<!-- ```{r, eval = FALSE} -->

<!-- # 1. Check for duplicate patients in the demographics dataset -->
<!-- demog_dupes <- df_demog_clean %>% -->
<!--   count(patient) %>% -->
<!--   filter(n > 1) -->

<!-- cat("Number of patients with multiple records in df_demog_clean:", nrow(demog_dupes), "\n") -->

<!-- # If you want to view the actual duplicate rows, uncomment the following line: -->
<!-- # df_demog_clean %>% filter(patient %in% demog_dupes$patient) %>% arrange(patient) -->


<!-- # 2. Check for duplicate patients in the 6-month visit dataset -->
<!-- visit_dupes <- df_visits_clean %>% -->
<!--   count(patient) %>% -->
<!--   filter(n > 1) -->

<!-- cat("Number of patients with multiple records in df_6m_visit:", nrow(visit_dupes), "\n") -->


<!-- # 3. Analyze patient overlap between the two datasets -->
<!-- demog_ids <- unique(df_demog_clean$patient) -->
<!-- visit_ids <- unique(df_visits_clean$patient) -->

<!-- # Patients in both datasets -->
<!-- overlap_ids <- intersect(demog_ids, visit_ids) -->
<!-- cat("Number of patients in BOTH datasets:", length(overlap_ids), "\n") -->

<!-- # Patients in demographics but missing from the 6-month visit -->
<!-- # (This represents dropouts, deaths, or incomplete follow-ups) -->
<!-- demog_only_ids <- setdiff(demog_ids, visit_ids) -->
<!-- cat("Number of patients in df_demog_clean ONLY:", length(demog_only_ids), "\n") -->

<!-- # Patients in the 6-month visit but missing from baseline demographics -->
<!-- # (Ideally, this should be 0; if not, there may be data entry issues) -->
<!-- visit_only_ids <- setdiff(visit_ids, demog_ids) -->
<!-- cat("Number of patients in df_visits_clean ONLY:", length(visit_only_ids), "\n") -->

<!-- ``` -->

